---
title: "Methylation–Expression Concordance: DMRs vs Differential Expression"
format:
  pdf:
    toc: true
    toc-depth: 2
    toc-title: "Table of Contents"
    number-sections: true
    keep-tex: false
    geometry:
      - margin=1in
    fontsize: 10pt
    papersize: letter
    fig-pos: "H"          # keep figures where they're placed
    tables: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(SummarizedExperiment)
library(GenomicRanges)
library(GenomeInfoDb)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(here)

# expression data
comparison_df <- read_csv(here("data/earli_limma_deseq_comp.csv"))
earli_se      <- readRDS(here("data/earli_summarizedexperiment.rds"))
# unfiltered dmr data
dmr_gr <- readRDS(here("data/private/DMRs.Granges.rds"))

row_meta <- rowRanges(earli_se) |>
    as.data.frame() |>
    as_tibble(rownames = "ensg") |>
    dplyr::select(ensg, chr = seqnames, start, end, strand,
                  gene_symbol = SYMBOL, gene_name = GENENAME,
                  gene_type = GENETYPE, autosomal_protein_coding,
                  hemoglobin_related)

# note: no significance filter
all_expr <- comparison_df |>
    left_join(row_meta, by = "ensg") |>
    filter(!is.na(gene_symbol), !is.na(deseq2_lfc))


# Classify each DMR as intragenic (overlaps a gene body) or intergenic.
# dist_to_gene is distance to nearest feature regardless of strand
gene_ref <- genes(TxDb.Hsapiens.UCSC.hg38.knownGene)
gene_ref <- keepSeqlevels(gene_ref,
                          intersect(seqlevels(dmr_gr), seqlevels(gene_ref)),
                          pruning.mode = "coarse")

d2n <- distanceToNearest(dmr_gr, gene_ref, ignore.strand = TRUE)

dmr_gr_annot <- as_tibble(dmr_gr) |>
    dplyr::rename(SYMBOL = symbol, dir = direction, dmr_p = p_value, dmr_fdr = p_fdr,
                  dmr_beta = mean_beta)

dmr_gr_annot$dist_to_gene <- NA_real_
dmr_gr_annot$dist_to_gene[queryHits(d2n)] <- mcols(d2n)$distance

dmr_gr_annot$dmr_class <- ifelse(dmr_gr_annot$dist_to_gene == 0,
                                  "intragenic", "intergenic")

# Where there are multiple DMR for a given gene, choose the one with the
# smallest pval
dmr_per_gene <- dmr_gr_annot |>
    group_by(SYMBOL) |>
    slice_min(dmr_p, n = 1, with_ties = FALSE) |>
    ungroup() |>
    dplyr::select(SYMBOL, dmr_beta, dmr_p, dmr_fdr, dir, dmr_class, dist_to_gene)

# All genes with a DMR and expression data; no significance filter on either side.
# deseq2_lfc > 0 → up in Hyper; dmr_beta > 0 → more methylated in Hyper.
# Methylation-suppresses-expression predicts opposite signs → negative Spearman ρ.
multi_omic <- all_expr |>
    inner_join(dmr_per_gene, by = c("gene_symbol" = "SYMBOL")) |>
    mutate(
        is_de = !is.na(deseq2_padj) & deseq2_padj < 0.05,
        methyl_concordant = (dmr_beta > 0 & deseq2_lfc < 0) |
                            (dmr_beta < 0 & deseq2_lfc > 0),
        quadrant = case_when(
            dmr_beta > 0 & deseq2_lfc < 0 ~ "More methyl in Hyper / Down in Hyper",
            dmr_beta < 0 & deseq2_lfc > 0 ~ "Less methyl in Hyper / Up in Hyper",
            dmr_beta > 0 & deseq2_lfc > 0 ~ "More methyl in Hyper / Up in Hyper",
            dmr_beta < 0 & deseq2_lfc < 0 ~ "Less methyl in Hyper / Down in Hyper"
        )
    )
```

```{r}
# write_csv(multi_omic, here("data/private/dmr_de_inner.csv"))
```


## Introduction

Differential methylated regions were discovered between the hyper and hypo
sepsis phenotypes. The question is how these regions relate to an entirely 
independent bulk RNAseq dataset made up of sepsis patients who are also labeled
as hyper or hypo inflammatory.

Differential expression was performed with both DESeq2 and limma voom since
there is some belief that DESeq2 has [inflated FDR in large population studies](https://pmc.ncbi.nlm.nih.gov/articles/PMC8922736/).
That doesn't seem to be the case for this data, and DESeq2 and limma voom nearly
entirely agree. I recommend using the deseq lfc and padj for analysis.

The hypothesis is that **methylation suppresses expression**. Thus, genes with
higher differential methylation in Hyper patients (positive `dmr_beta`) should
have lower expression in Hyper relative to Hypo (negative or near-zero log2FC),
and vice versa. This predicts a **negative** Spearman correlation between DMR
mean beta and DESeq2 log2FC.


## Data dictionary

This is the dataframe shared with this notebook, which is an inner join
(only gene symbols in both the unfiltered DE results and unfiltered DMR
results) between the unfiltered differential expression results (both
DESeq2 and limma-voom), and the unfiltered DMR regions (from the genomic
ranges obj). For the DE results, I recommend using the deseq lfc and padj.
The results of the two are very nearly identical, however.

```{r data-dictionary}
dict <- data.frame(
  Group = c(
    rep("Gene identity", 8),
    rep("DESeq2 (Hyper vs Hypo)", 4),
    rep("limma/voom (Hyper vs Hypo)", 4),
    rep("Method concordance", 2),
    rep("DMR", 6),
    rep("Analysis flags", 3)
  ),
  Column = c(
    "ensg", "gene_symbol", "gene_name", "gene_type",
    "chr / start / end", "strand",
    "autosomal_protein_coding", "hemoglobin_related",
    "deseq2_lfc", "deseq2_pval", "deseq2_padj", "sig_deseq2",
    "limma_lfc", "limma_pval", "limma_padj", "sig_limma",
    "concordance", "dir_concordance",
    "dmr_beta", "dmr_p", "dmr_fdr", "dir", "dmr_class", "dist_to_gene",
    "is_de", "methyl_concordant", "quadrant"
  ),
  Description = c(
    "Ensembl gene ID (primary key)",
    "HGNC gene symbol",
    "Full gene name",
    'Biotype (e.g. "protein-coding")',
    "Genomic coordinates of the gene body (hg38)",
    "Transcription strand (+ / -)",
    "TRUE if the gene is autosomal and protein-coding",
    "TRUE if the gene encodes a hemoglobin subunit",
    "Log2 fold-change (Hyper / Hypo); positive = higher in Hyper. Wald test on MLE (un-shrunken) estimates.",
    "Wald test p-value",
    "BH-adjusted p-value",
    "TRUE if deseq2_padj < 0.05",
    "Log2 fold-change (Hyper / Hypo); positive = higher in Hyper.",
    "Moderated t-test p-value",
    "BH-adjusted p-value",
    "TRUE if limma_padj < 0.05",
    '"Both" / "DESeq2 only" / "limma only" / "Neither" based on padj < 0.05 in each method',
    '"Both (up)" / "Both (down)" / other concordance levels',
    "Mean beta-value difference at the DMR (Hyper \u2212 Hypo); positive = more methylated in Hyper",
    "Smallest DMR p-value across all DMRs mapped to this gene",
    "BH-adjusted DMR p-value",
    '"hyper" = more methylated in Hyper; "hypo" = more methylated in Hypo',
    '"intragenic" if the DMR overlaps the gene body; "intergenic" otherwise',
    "Distance in bp from the DMR to the nearest gene body (TxDb hg38); 0 for intragenic DMRs",
    "TRUE if deseq2_padj < 0.05",
    "TRUE if dmr_beta and deseq2_lfc have opposite signs, consistent with methylation-suppresses-expression",
    'Quadrant label combining DMR and expression direction, e.g. "More methyl in Hyper / Down in Hyper"'
  ),
  stringsAsFactors = FALSE
)

knitr::kable(dict, align = c("l", "l", "l"))
```

## Overview

```{r overlap-summary}
n_expr <- n_distinct(all_expr$gene_symbol)
n_dmr  <- n_distinct(dmr_gr_annot$SYMBOL)
n_ovlp <- n_distinct(multi_omic$gene_symbol)

knitr::kable(
    data.frame(
        Category = c(
            "Genes tested for DE (DESeq2)",
            "Genes with ≥1 DMR (unfiltered)",
            "Overlap: genes with both expression and DMR data"
        ),
        N = scales::comma(c(n_expr, n_dmr, n_ovlp))
    ),
    col.names = c("Category", "N"),
    align = c("l", "r")
)
```

## Methylation vs expression

```{r dmr-de-scatter-quadrant}
#| fig-width: 14
#| fig-height: 6

cor_test <- suppressWarnings(
    cor.test(multi_omic$dmr_beta, multi_omic$deseq2_lfc, method = "spearman")
)
rho <- cor_test$estimate

p_scatter <- ggplot(multi_omic,
       aes(x = dmr_beta, y = deseq2_lfc,
           color = is_de, alpha = is_de, size = is_de)) +
    annotate("rect", xmin =    0, xmax =  Inf, ymin = -Inf, ymax = 0,
             fill = "steelblue", alpha = 0.3) +
    annotate("rect", xmin = -Inf, xmax =    0, ymin =    0, ymax = Inf,
             fill = "steelblue", alpha = 0.3) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_point() +
    scale_color_manual(
        values = c("TRUE" = "firebrick", "FALSE" = "grey60"),
        labels = c("TRUE" = "DE (padj < 0.05)", "FALSE" = "Not DE")
    ) +
    scale_alpha_manual(values = c("TRUE" = 0.8, "FALSE" = 0.3), guide = "none") +
    scale_size_manual(values  = c("TRUE" = 1.8, "FALSE" = 1.0), guide = "none") +
    labs(
        x        = "DMR mean beta (positive = more methylated in Hyper)",
        y        = "DESeq2 log2FC (positive = higher expression in Hyper)",
        color    = NULL,
        title    = "Methylation vs expression: all DMR genes",
        subtitle = sprintf(
            "Spearman \u03c1 = %.3f, p = %s | n = %d",
            rho, signif(cor_test$p.value, 2), nrow(multi_omic)
        )
    ) +
    theme_bw(base_size = 15) +
    theme(legend.position = "bottom")

p_quad <- multi_omic |>
    dplyr::count(quadrant, methyl_concordant) |>
    mutate(quadrant = str_wrap(quadrant, width = 30)) |>
    ggplot(aes(y = reorder(quadrant, n), x = n, fill = methyl_concordant)) +
    geom_col() +
    geom_text(aes(label = scales::comma(n)), hjust = -0.2, size = 5) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.15)),
                       labels = scales::comma) +
    scale_fill_manual(
        values = c("TRUE" = "steelblue", "FALSE" = "firebrick"),
        labels = c("TRUE" = "Expected (methyl \u2192 suppress)", "FALSE" = "Unexpected")
    ) +
    labs(
        y        = NULL,
        x        = "Number of genes",
        fill     = NULL,
        title    = "Quadrant counts",
        subtitle = "Blue = expected under methyl \u2192 suppress"
    ) +
    theme_bw(base_size = 15) +
    theme(legend.position = "bottom")

gridExtra::grid.arrange(p_scatter, p_quad, ncol = 2, widths = c(1.3, 1))
```

A positive correlation means that more methylation is correlated with greater
expression, the opposite of our hypothesis.

## DMR region classification

Each of the `r nrow(dmr_gr_annot)` DMRs was classified using distance to the
nearest gene body (TxDb hg38, `r scales::comma(length(gene_ref))` genes).
- **intragenic** if the DMR overlaps a gene body (distance = 0)
- **intergenic** otherwise.

```{r dmr-class-dist}
#| fig-width: 8
#| fig-height: 4

intergenic <- dmr_gr_annot[dmr_gr_annot$dmr_class == "intergenic" &
                            !is.na(dmr_gr_annot$dist_to_gene), ]
med_dist <- median(intergenic$dist_to_gene)

p_bar <- dmr_gr_annot |>
    dplyr::count(dmr_class) |>
    ggplot(aes(x = dmr_class, y = n, fill = dmr_class)) +
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = scales::comma(n)), vjust = -0.4, size = 4) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.12)),
                       labels = scales::comma) +
    scale_fill_manual(values = c("intragenic" = "steelblue",
                                 "intergenic" = "goldenrod")) +
    labs(x = NULL, y = "Number of DMRs", title = "DMR class counts") +
    theme_bw(base_size = 13)

p_hist <- ggplot(intergenic, aes(x = log10(dist_to_gene))) +
    geom_histogram(bins = 40, fill = "goldenrod", color = "white") +
    geom_vline(xintercept = log10(med_dist), color = "firebrick",
               linewidth = 1, linetype = "dashed") +
    annotate("text", x = log10(med_dist) + 0.05,
             y = Inf, vjust = 1.5, hjust = 0, size = 3.5, color = "firebrick",
             label = sprintf("median = %s bp", scales::comma(round(med_dist)))) +
    labs(x = "log10(distance to nearest gene, bp)", y = "Count",
         title = "Intergenic DMR distances") +
    theme_bw(base_size = 13)

gridExtra::grid.arrange(p_bar, p_hist, ncol = 2)
```


```{r dmr-class-correlation}
cor_by_class <- multi_omic |>
    group_by(dmr_class) |>
    summarise(
        n   = n(),
        rho = suppressWarnings(
            cor(dmr_beta, deseq2_lfc, method = "spearman")
        ),
        pval = suppressWarnings(
            cor.test(dmr_beta, deseq2_lfc, method = "spearman")$p.value
        ),
        pct_expected = round(100 * mean(methyl_concordant), 1),
        .groups = "drop"
    )

knitr::kable(
    cor_by_class |>
        dplyr::select(
            `DMR class`    = dmr_class,
            N              = n,
            `Spearman rho` = rho,
            `p-value`      = pval,
            `% expected`   = pct_expected
        ) |>
        mutate(across(`Spearman rho`, round, 3),
               across(`p-value`, \(x) signif(x, 3))),
    caption = "Methylation-expression correlation by DMR class"
)
```

```{r dmr-class-scatter}
#| fig-width: 10
#| fig-height: 5

multi_omic |>
    mutate(dmr_class = factor(dmr_class,
                              levels = c("intragenic", "intergenic"),
                              labels = c("Intragenic", "Intergenic"))) |>
    ggplot(aes(x = dmr_beta, y = deseq2_lfc,
               color = is_de, alpha = is_de, size = is_de)) +
    annotate("rect", xmin =    0, xmax =  Inf, ymin = -Inf, ymax = 0,
             fill = "steelblue", alpha = 0.3) +
    annotate("rect", xmin = -Inf, xmax =    0, ymin =    0, ymax = Inf,
             fill = "steelblue", alpha = 0.3) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_point() +
    scale_color_manual(
        values = c("TRUE" = "firebrick", "FALSE" = "grey60"),
        labels = c("TRUE" = "DE (padj < 0.05)", "FALSE" = "Not DE")
    ) +
    scale_alpha_manual(values = c("TRUE" = 0.8, "FALSE" = 0.3), guide = "none") +
    scale_size_manual(values  = c("TRUE" = 1.8, "FALSE" = 1.0), guide = "none") +
    facet_wrap(~dmr_class) +
    labs(
        x     = "DMR mean beta (positive = more methylated in Hyper)",
        y     = "DESeq2 log2FC (positive = higher expression in Hyper)",
        color = NULL,
        title = "Methylation vs expression by DMR class",
        subtitle = "Shaded quadrants = expected under methyl \u2192 suppress"
    ) +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")
```

Neither class shows a meaningful expected signal.

## DMRs near the gene body show greater expected directional effect on expression

```{r dmr-distance-gradient}
#| fig-width: 8
#| fig-height: 5

intergenic_mo <- multi_omic |>
    filter(dmr_class == "intergenic") |>
    mutate(
        dist_bin = cut(dist_to_gene,
                       breaks = quantile(dist_to_gene,
                                         probs = seq(0, 1, by = 0.25)),
                       include.lowest = TRUE,
                       labels = c("Q1\n(nearest)", "Q2", "Q3", "Q4\n(farthest)"))
    )

dist_cor <- intergenic_mo |>
    group_by(dist_bin) |>
    summarise(
        n    = n(),
        rho  = suppressWarnings(cor(dmr_beta, deseq2_lfc, method = "spearman")),
        pval = suppressWarnings(
            cor.test(dmr_beta, deseq2_lfc, method = "spearman")$p.value),
        dist_median = median(dist_to_gene),
        .groups = "drop"
    )

ggplot(dist_cor, aes(x = dist_bin, y = rho)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_col(fill = "goldenrod", width = 0.6) +
    geom_text(aes(label = sprintf("p = %s\nn = %d", signif(pval, 2), n)),
              vjust = ifelse(dist_cor$rho >= 0, -0.3, 1.2), size = 3.5) +
    scale_y_continuous(expand = expansion(mult = c(0.15, 0.15))) +
    labs(
        x        = "Distance to nearest gene (by quartile)",
        y        = "Spearman \u03c1 (dmr_beta vs log2FC)",
        title    = "Methylation\u2013expression correlation by distance from gene body",
        subtitle = "Intergenic DMRs only | negative = expected under methyl \u2192 suppress"
    ) +
    theme_bw(base_size = 13)
```

```{r dmr-distance-table}
knitr::kable(
    dist_cor |>
        dplyr::select(
            `Distance quartile`  = dist_bin,
            `Median distance (bp)` = dist_median,
            N = n,
            `Spearman rho`       = rho,
            `p-value`            = pval
        ) |>
        mutate(
            `Median distance (bp)` = scales::comma(round(`Median distance (bp)`)),
            `Spearman rho`         = round(`Spearman rho`, 3),
            `p-value`              = signif(`p-value`, 3)
        ),
    caption = "Methylation-expression correlation by intergenic distance quartile"
)
```

The two nearest quartiles (Q1, median
`r scales::comma(round(dist_cor$dist_median[1]))` bp; Q2, median
`r scales::comma(round(dist_cor$dist_median[2]))` bp, max ~12,300 bp) both show
slightly negative ρ, which is the expected direction given the assumption that 
higher methylation depresses expression, while Q3 and Q4 flip positive, with
the farthest DMRs showing the strongest
positive ρ (ρ = `r round(dist_cor$rho[4], 3)`, p = `r signif(dist_cor$pval[4], 2)`).
