---
title: "EARLI LCA Differential Expression"
format:
  pdf:
    toc: true
    toc-depth: 2
    toc-title: "Table of Contents"
    number-sections: true
    keep-tex: false
    geometry:
      - margin=1in
    fontsize: 10pt
    papersize: letter
    fig-pos: "H"          # keep figures where they're placed
    tables: true
execute:
  echo: false
  warning: false
  message: false
  cache: true
  cache-lazy: false
  fig-width: 9
  fig-height: 6
---

```{r setup}
library(GEOquery)
library(limma)
library(tidyverse)
library(DESeq2)
library(here)
library(sva)
library(edgeR)
library(RUVSeq)
library(matrixStats)
library(parallel)
library(pheatmap)
library(ggrepel)
library(GenomicRanges)

theme_set(theme_bw(base_size = 14))
```

## read in EARLI summarized experiment

see r/earli_se.r

```{r}
earli_se = readRDS(here("data/earli_summarizedexperiment.rds"))
```


## Count-level EDA

### Library sizes

```{r library-sizes}
#| fig-height: 9
lib_df <- tibble(
    sample       = colnames(earli_se),
    total_counts = colSums(assay(earli_se, "counts")),
    lca_label    = earli_se$lca_label
)

ggplot(lib_df, aes(x = reorder(sample, total_counts), y = total_counts / 1e6, fill = lca_label)) +
    geom_col() +
    coord_flip() +
    labs(x = NULL, y = "Total counts (millions)", title = "Library sizes by sample")
```

Library sizes range from 4.0 M to 44.1 M reads (median 10.4 M, n = 189).
Most samples fall between 8 M and 15 M reads. One Hyper sample (EARLI_11533)
and two neighboring samples exceed 40 M reads and stand out visually.

Is there any relationship between LCA label and library size?

```{r library-size-by-lca}
wt <- wilcox.test(total_counts ~ lca_label, data = lib_df)

lib_df |>
    ggplot(aes(x = lca_label, y = total_counts / 1e6, fill = lca_label)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    labs(
        x        = NULL,
        y        = "Total counts (millions)",
        title    = "Library size by LCA label",
        subtitle = paste0("Wilcoxon p = ", signif(wt$p.value, 3))
    )
```

Library depth does not differ meaningfully between subgroups
(Hyper median 10.0 M, Hypo median 10.5 M; Wilcoxon p = 0.769).
Library size is unlikely to confound the primary Hyper vs. Hypo contrast.

### Top genes by library fraction

Are any genes dominating individual libraries?

```{r library-composition}
cnt_totals <- rowSums(assay(earli_se, "counts"))
lib_total  <- sum(cnt_totals)

tibble(ensg = names(cnt_totals), lib_pct = cnt_totals / lib_total * 100) |>
    left_join(
        as_tibble(mcols(rowRanges(earli_se))) |>
            mutate(ensg = names(rowRanges(earli_se))),
        by = "ensg"
    ) |>
    arrange(desc(lib_pct)) |>
    mutate(cumulative_pct = cumsum(lib_pct)) |>
    slice_head(n = 30) |>
    ggplot(aes(x = reorder(SYMBOL, lib_pct), y = lib_pct)) +
    geom_col() +
    coord_flip() +
    labs(x = NULL, y = "% of total library", title = "Top 30 genes by library fraction")
```

No single gene dominates library composition. *S100A9* contributes the
largest fraction (1.6% of total counts), with the top five genes
(*S100A9*, *PCSK5*, *B2M*, *ACTB*, *S100A8*) accounting for ~4.5% of
reads combined. Not concerning.

### PCA

```{r pca}
vsd <- vst(DESeqDataSet(earli_se, design = ~ age_scaled + sex + lca_label), blind = TRUE)

pca_res <- prcomp(t(assay(vsd)), scale. = FALSE)
pct_var <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

tibble(PC = seq_along(pct_var), pct_var = pct_var) |>
    slice_head(n = 20) |>
    ggplot(aes(x = PC, y = pct_var)) +
    geom_col() + geom_line() + geom_point() +
    labs(y = "% variance explained", title = "Scree plot")
```

PC1 explains 19.8% of variance and PC2 explains 9.4%, with a pronounced
drop-off after the first components and a slower decline thereafter.
The first four PCs together account for ~42% of total variance.

```{r pca-plots-setup}
col_data <- as.data.frame(colData(earli_se))

pca_df <- as_tibble(pca_res$x[, 1:4], rownames = "sample") |>
    left_join(as_tibble(col_data, rownames = "sample"), by = "sample")

pc_labs <- function(i, j) {
    labs(
        x = paste0("PC", i, " (", pct_var[i], "%)"),
        y = paste0("PC", j, " (", pct_var[j], "%)")
    )
}
```

```{r pca-plots}
tibble(PC = seq_along(pct_var), pct_var = pct_var) |>
    slice_head(n = 20) |>
    ggplot(aes(x = PC, y = pct_var)) +
    geom_col() + geom_line() + geom_point() +
    labs(y = "% variance explained", title = "Scree plot")

ggplot(pca_df, aes(PC1, PC2, color = lca_label)) +
    geom_point() + pc_labs(1, 2) + ggtitle("PC1 vs PC2 — LCA label")

ggplot(pca_df, aes(PC1, PC2, color = sex)) +
    geom_point() + pc_labs(1, 2) + ggtitle("PC1 vs PC2 — sex")

ggplot(pca_df, aes(PC1, PC2, color = age_scaled)) +
    geom_point() +
    scale_color_viridis_c() +
    pc_labs(1, 2) + ggtitle("PC1 vs PC2 — age (scaled)")
```

LCA label and sex do not produce a clear separation in PC1/PC2 space. 
The two groups overlap substantially. Three samples are notable outliers
along PC1 (PC1 > 70), spanning both LCA subgroups, suggesting their
separation in this component is not LCA-driven. No strong age gradient is
visible in the first two PCs.

Which PC is most associated with LCA label?

```{r pc-lca-association}
n_pcs <- min(20, ncol(pca_res$x))

pc_lca_df <- as_tibble(pca_res$x[, 1:n_pcs], rownames = "sample") |>
    left_join(as_tibble(col_data, rownames = "sample"), by = "sample")

pc_assoc <- tibble(
    pc     = paste0("PC", 1:n_pcs),
    pc_idx = 1:n_pcs,
    p_val  = map_dbl(1:n_pcs, \(i) {
        kruskal.test(pca_res$x[, i] ~ earli_se$lca_label)$p.value
    })
) |>
    arrange(p_val) |>
    mutate(p_adj = p.adjust(p_val, method = "BH"))

print(pc_assoc)

best_pc <- pc_assoc$pc[1]
cat("\nPC most associated with LCA label:", best_pc,
    sprintf("(Kruskal-Wallis p = %.3g)", pc_assoc$p_val[1]), "\n")

ggplot(pc_lca_df, aes(lca_label, .data[[best_pc]], fill = lca_label)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    geom_jitter(width = 0.15, size = 1) +
    labs(
        x     = "LCA label",
        y     = paste0(best_pc, " (", pct_var[pc_assoc$pc_idx[1]], "% variance)"),
        title = paste(best_pc, "~ LCA label (most associated PC)")
    ) +
    theme(legend.position = "none")
```

PC4 (6.1% of variance) is the PC most strongly associated with LCA label
(Kruskal-Wallis p = 1.3e-10, BH-adjusted p = 2.6e-9), followed by
PC2 (p = 3.1e-7). The primary biological signal thus sits in PCs 2 and 4
rather than the top component, which likely captures technical or other
sources of variation dominant in this cohort.

## Latent factor estimation

Both SVAseq and RUVseq (RUVr) are run to infer unknown sources of variation
while protecting the `lca_label` effect of interest. SVAseq uses a full vs.
null model contrast to isolate variance unexplained by the biology of interest.
RUVr uses residuals from a first-pass model that includes `lca_label`, so the
factors it finds are orthogonal to the primary contrast by construction.

### SVAseq

```{r sva}
dds_norm      <- estimateSizeFactors(DESeqDataSet(earli_se, design = ~ age_scaled + sex + lca_label))
norm_log_cnts <- log1p(counts(dds_norm, normalized = TRUE))

mod  <- model.matrix(~ age_scaled + sex + lca_label, data = col_data)
mod0 <- model.matrix(~ age_scaled + sex,             data = col_data)

n_sv <- num.sv(norm_log_cnts, mod)
cat("Estimated number of SVs:", n_sv, "\n")

sva_res <- svaseq(norm_log_cnts, mod, mod0, n.sv = n_sv)
```

SVAseq estimates **15 surrogate variables**, indicating a substantial
amount of variance in the data that is unexplained by
`lca_label`, `age_scaled`, or `sex`.

### RUVseq (RUVr)

```{r ruv}
design_ruv <- model.matrix(~ lca_label + age_scaled + sex, data = col_data)

y <- DGEList(counts = assay(earli_se, "counts"))
y <- calcNormFactors(y, method = "upperquartile")
y <- estimateGLMCommonDisp(y, design_ruv)
y <- estimateGLMTagwiseDisp(y, design_ruv)
fit <- glmFit(y, design_ruv)
res <- residuals(fit, type = "deviance")

set     <- newSeqExpressionSet(assay(earli_se, "counts"))
ruv_res <- RUVr(set, rownames(set), k = n_sv, residuals = res)
```

### RUV k selection

`k` is initialized from SVA's `num.sv()` estimate. Two diagnostics assess
whether that choice is appropriate for RUVr:

1. **Scree plot of deviance residuals** — singular values of the residual
   matrix fed to RUVr. An elbow indicates a natural cutoff.
2. **Genomic inflation (λ) vs k** — for each candidate k, RUVr is re-run
   and edgeR LRT p-values for `lca_label` are extracted.
   λ = median(x^2) / expected median under null; values near 1 indicate
   well-calibrated tests.

```{r ruv-scree}
sv_decomp <- svd(res)
sv_pct    <- sv_decomp$d^2 / sum(sv_decomp$d^2) * 100

tibble(k = seq_along(sv_pct), pct_var = sv_pct) |>
    slice_head(n = 30) |>
    ggplot(aes(x = k, y = pct_var)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    geom_line() + geom_point() +
    geom_vline(xintercept = n_sv, linetype = "dashed", color = "firebrick") +
    annotate("text", x = n_sv + 0.4, y = sv_pct[1] * 0.9,
             label = paste0("SVA k = ", n_sv), hjust = 0, color = "firebrick") +
    labs(
        x        = "Component (k)",
        y        = "% variance in deviance residuals",
        title    = "Scree plot of RUVr input (deviance residuals)",
        subtitle = "Dashed line = k chosen by SVAseq num.sv()"
    )
```

The first deviance-residual component captures 23.0% of residual variance,
with a clear drop after the first component and a plateau beginning
around k = 6. The SVAseq-suggested k = 15 captures 60.5% of cumulative
residual variance.

```{r ruv-k-selection}
k_max <- min(n_sv + 5L, ncol(assay(earli_se, "counts")) - 1L)

Y_svd <- svd(t(res), nu = k_max, nv = 0)

y_base <- DGEList(counts = assay(earli_se, "counts"))
y_base <- calcNormFactors(y_base, method = "upperquartile")
y_base <- estimateGLMCommonDisp(y_base, design_ruv)
y_base <- estimateGLMTagwiseDisp(y_base, design_ruv)

lambda_df <- mclapply(seq_len(k_max), function(k) {
    W <- Y_svd$u[, 1:k, drop = FALSE]
    design_k <- cbind(
        model.matrix(~ lca_label + age_scaled + sex, data = col_data),
        W
    )
    fit_k  <- glmFit(y_base, design_k)
    lrt_k  <- glmLRT(fit_k, coef = "lca_labelHyper")
    pvals  <- lrt_k$table$PValue
    chisq  <- qchisq(1 - pvals, df = 1)
    lambda <- median(chisq, na.rm = TRUE) / qchisq(0.5, df = 1)
    n_sig  <- sum(p.adjust(pvals, method = "BH") < 0.05, na.rm = TRUE)
    tibble(k = k, lambda = lambda, n_sig = n_sig)
}, mc.cores = 20L) |>
    bind_rows()
```

```{r ruv-lambda-plot}
lambda_df |>
    ggplot(aes(x = k, y = lambda)) +
    geom_line() + geom_point() +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
    geom_vline(xintercept = n_sv, linetype = "dashed", color = "firebrick") +
    annotate("text", x = n_sv + 0.3, y = max(lambda_df$lambda),
             label = paste0("SVA k = ", n_sv), hjust = 0, color = "firebrick") +
    labs(
        x        = "k (RUV factors)",
        y        = "Genomic inflation (λ)",
        title    = "Genomic inflation vs RUV k",
        subtitle = "λ ≈ 1 = well-calibrated | λ > 1 = inflation | λ < 1 = deflation"
    )

lambda_df |>
    ggplot(aes(x = k, y = n_sig)) +
    geom_line() + geom_point() +
    geom_vline(xintercept = n_sv, linetype = "dashed", color = "firebrick") +
    labs(x = "k (RUV factors)", y = "# genes BH < 0.05",
         title = "Significant DE genes vs RUV k")
```

Genomic inflation (λ) remains between 4.4 and 4.9 across all k values
tested, well above the conventional threshold of λ ~1, and does not
meaningfully decrease as k increases. This is probably b/c the hyper and 
hypo group are very different, and hence there are a lot of highly significant 
DE genes across the range of tested k. The number of significant DE genes
stabilizes between ~5,000 and ~5,400 from k = 13 onward, supporting the sva 
inferred k = 15 as a reasonable choice.

### RLE plots

RLE (relative log expression) plots show per-sample distributions of expression
values after subtracting each gene's median across all samples. Well-corrected
data has boxes centered near 0 with similar widths across samples.

```{r ruv-rle}
#| fig-height: 8
set.seed(42)
gene_idx     <- sample(nrow(norm_log_cnts), min(5000L, nrow(norm_log_cnts)))
k_candidates <- sort(unique(c(0L, 1L, floor(k_max / 3L), floor(2L * k_max / 3L), n_sv, k_max)))

rle_df <- map_dfr(k_candidates, function(k) {
    expr <- if (k == 0L) {
        norm_log_cnts[gene_idx, ]
    } else {
        W <- Y_svd$u[, 1:k, drop = FALSE]
        removeBatchEffect(norm_log_cnts[gene_idx, ], covariates = W)
    }
    row_meds <- apply(expr, 1, median)
    rle_mat  <- expr - row_meds

    as_tibble(t(rle_mat), rownames = "sample") |>
        pivot_longer(-sample, names_to = "gene", values_to = "rle") |>
        mutate(k = factor(paste0("k = ", k), levels = paste0("k = ", k_candidates)))
})

rle_df |>
    ggplot(aes(x = sample, y = rle)) +
    geom_boxplot(outlier.shape = NA, linewidth = 0.3) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "firebrick", linewidth = 0.4) +
    coord_cartesian(ylim = c(-2, 2)) +
    facet_wrap(~ k, ncol = 1) +
    theme_bw() +
    theme(axis.text.x = element_blank()) +
    labs(
        x        = NULL,
        y        = "Relative log expression",
        title    = "RLE plots across candidate RUV k values",
        subtitle = "Well-corrected: boxes centered near 0 with uniform widths"
    )

rle_df |>
    group_by(sample, k) |>
    reframe(iqr = IQR(rle), median_dev = abs(median(rle))) |>
    pivot_longer(c(iqr, median_dev), names_to = "metric", values_to = "value") |>
    mutate(metric = recode(metric,
        iqr        = "IQR (spread)",
        median_dev = "|median| (centering)"
    )) |>
    ggplot(aes(x = k, y = value)) +
    geom_boxplot() +
    facet_wrap(~ metric, scales = "free_y") +
    theme_bw() +
    labs(
        x        = "k (RUV factors)",
        y        = NULL,
        title    = "RLE summary across samples by k",
        subtitle = "Lower IQR = less spread; lower |median| = better centering near 0"
    )
```

Mean per-sample IQR decreases from 0.73 at k = 0 to 0.46 at k = 15, and median deviation (centering near zero) falls from 0.062 to 0.013, indicating steady improvement in between-sample comparability. The gains diminish noticeably beyond k ≈ 6–13, suggesting k = 15 provides solid correction without substantial over-fitting.

### Factor characterization

Spearman correlations between each latent factor and known sample variables.
A strong correlation with a known variable suggests that factor is capturing
that source of variation rather than a truly unknown confounder.

```{r factor-characterization}
known_vars <- tibble(
    lib_size  = colSums(assay(earli_se, "counts")),
    age       = earli_se$age_scaled,
    sex       = as.numeric(earli_se$sex == "Male"),
    lca_label = as.numeric(earli_se$lca_label == "Hyper")
)

sv_mat <- sva_res$sv
colnames(sv_mat) <- paste0("SV", seq_len(ncol(sv_mat)))

ruv_mat <- pData(ruv_res)[, grep("W_", colnames(pData(ruv_res))), drop = FALSE]
colnames(ruv_mat) <- paste0("RUV", seq_len(ncol(ruv_mat)))

all_vars <- cbind(as.matrix(known_vars), sv_mat, as.matrix(ruv_mat))
cor_mat  <- cor(all_vars, method = "spearman")
cor_sub  <- cor_mat[colnames(known_vars), c(colnames(sv_mat), colnames(ruv_mat))]

pheatmap(
    cor_sub,
    main            = "Spearman correlation: known variables vs latent factors",
    color           = colorRampPalette(c("blue", "white", "red"))(50),
    breaks          = seq(-1, 1, length.out = 51),
    display_numbers = TRUE,
    number_format   = "%.2f",
    cluster_cols    = FALSE
)
```

SV6 shows the strongest correlation with any known variable (Spearman ρ = 0.81 with sex), indicating it primarily captures sex-related expression differences. SV3 correlates moderately with library size (ρ = −0.42). Most RUV factors show only weak correlations with all known covariates — as expected, given they are derived from residuals of a model that already conditions on `lca_label`, `age_scaled`, and `sex`.

## Differential expression: Hyper vs Hypo

Three RUV factors (k = 3) are included as covariates in both models. The W
factors are orthogonal to `lca_label`, `age_scaled`, and `sex` by construction
(derived from residuals of a model containing all three), so including them
does not introduce collinearity with the primary contrast.

```{r de-setup}
k_de <- 3L
W_de <- Y_svd$u[, 1:k_de, drop = FALSE]
colnames(W_de) <- paste0("W", seq_len(k_de))

col_data     <- as.data.frame(colData(earli_se))
col_data_ruv <- col_data
col_data_ruv[, colnames(W_de)] <- W_de

cnt_mat <- assay(earli_se, "counts")
```

---

### DESeq2

DESeq2 fits a negative binomial GLM per gene. For large observational studies
it can produce anti-conservative FDR estimates; results are presented alongside
limma/voom for comparison. `apeglm` shrinkage is applied to LFC for
visualization only — significance calls use the unshrunken Wald test p-values.

```{r deseq2-fit}
dds_ruv <- DESeqDataSetFromMatrix(
    countData = cnt_mat,
    colData   = col_data_ruv,
    design    = ~ age_scaled + sex + W1 + W2 + W3 + lca_label
)

# Filter low-count genes: CPM > 1 in at least the size of the smaller group
min_n   <- min(table(col_data_ruv$lca_label))
keep    <- rowSums(cpm(cnt_mat) > 1) >= min_n
dds_ruv <- dds_ruv[keep, ]

dds_ruv <- DESeq(dds_ruv)
```

```{r deseq2-results}
res_deseq2 <- results(
    dds_ruv,
    contrast = c("lca_label", "Hyper", "Hypo"),
    alpha    = 0.05
)

res_deseq2_shrunk <- lfcShrink(
    dds_ruv,
    coef = "lca_label_Hyper_vs_Hypo",
    type = "apeglm"
)

summary(res_deseq2)
```

DESeq2 identifies **7,275 DE genes** at BH-adjusted p < 0.05 from
16,457 tested: 3,905 (24%) upregulated in Hyper and 3,370 (20%) upregulated in
Hypo. No Cook's distance outlier genes were flagged.

```{r deseq2-pval-hist}
tibble(pvalue = res_deseq2$pvalue) |>
    filter(!is.na(pvalue)) |>
    ggplot(aes(x = pvalue)) +
    geom_histogram(binwidth = 0.02, boundary = 0,
                   fill = "steelblue", color = "white", alpha = 0.8) +
    geom_hline(
        yintercept = sum(!is.na(res_deseq2$pvalue)) * 0.02,
        linetype = "dashed", color = "firebrick"
    ) +
    labs(
        x        = "p-value",
        y        = "Count",
        title    = "DESeq2 p-value distribution: Hyper vs Hypo",
        subtitle = "Dashed line = expected count per bin under global null"
    ) +
    theme_bw()
```

The p-value histogram shows a pronounced spike near zero with counts well
above the null expectation (dashed line), consistent with widespread true
signal. The relatively flat tail suggests good test calibration outside the
region of signal.

```{r deseq2-plot-data}
deseq2_plot_df <- as_tibble(res_deseq2, rownames = "ensg") |>
    dplyr::select(ensg, baseMean, pvalue, padj) |>
    inner_join(
        as_tibble(res_deseq2_shrunk, rownames = "ensg") |>
            dplyr::select(ensg, log2FoldChange),
        by = "ensg"
    ) |>
    filter(!is.na(padj)) |>
    mutate(label = case_when(
        padj < 0.05 & log2FoldChange > 0 ~ "Up in Hyper",
        padj < 0.05 & log2FoldChange < 0 ~ "Up in Hypo",
        TRUE                              ~ "NS"
    ))
```

```{r deseq2-volcano}
deseq2_plot_df |>
    ggplot(aes(x = log2FoldChange, y = -log10(padj), color = label)) +
    geom_point(alpha = 0.4, size = 0.8) +
    scale_color_manual(values = c(
        "Up in Hyper" = "firebrick",
        "Up in Hypo"  = "steelblue",
        "NS"          = "grey70"
    )) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    labs(
        x        = "shrunken log2 fold change (Hyper / Hypo)",
        y        = "-log10(adjusted p-value)",
        color    = NULL,
        title    = "DESeq2 volcano: Hyper vs Hypo",
        subtitle = "apeglm-shrunken LFC | significance from unshrunken padj < 0.05"
    ) +
    theme_bw()
```

Hyper-upregulated genes reach larger shrunken fold changes (LFC up to ~4)
than Hypo-upregulated genes (LFC down to ~−2), suggesting stronger or more
consistent expression changes in the Hyper subgroup.
Several genes exceed −log10(padj) = 40.

```{r deseq2-ma}
deseq2_plot_df |>
    ggplot(aes(x = log10(baseMean), y = log2FoldChange, color = label)) +
    geom_point(alpha = 0.4, size = 0.8) +
    scale_color_manual(values = c(
        "Up in Hyper" = "firebrick",
        "Up in Hypo"  = "steelblue",
        "NS"          = "grey70"
    )) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    labs(
        x        = "log10(mean normalized count)",
        y        = "shrunken log2 fold change (Hyper / Hypo)",
        color    = NULL,
        title    = "DESeq2 MA plot: Hyper vs Hypo",
        subtitle = "apeglm shrinkage pulls low-count genes toward 0"
    ) +
    theme_bw()
```

apeglm shrinkage visibly compresses LFC estimates for low-abundance genes
toward zero, producing a narrowing funnel at low mean counts while
preserving large estimates for highly expressed genes.

---

### limma/voom

limma/voom uses a precision-weighted linear model on log-CPM values. It is
generally better calibrated than DESeq2 in large observational studies because
its empirical Bayes variance shrinkage is less sensitive to the mean-variance
relationship assumptions that can break down at large n. Two voom passes are
used: the first estimates the mean-variance trend, the second incorporates it
into precision weights for fitting.

```{r limma-fit}
dge <- DGEList(counts = cnt_mat[keep, ], group = col_data_ruv$lca_label)
dge <- calcNormFactors(dge, method = "TMM")

design_voom <- model.matrix(
    ~ age_scaled + sex + W1 + W2 + W3 + lca_label,
    data = col_data_ruv
)

# First voom pass: estimate mean-variance trend
v0     <- voom(dge, design_voom, plot = FALSE)
# Second voom pass: reweight with trend
v      <- voom(dge, design_voom, plot = TRUE)

fit_v  <- lmFit(v, design_voom)
fit_v  <- eBayes(fit_v, robust = TRUE)
```

```{r limma-results}
res_limma <- topTable(
    fit_v,
    coef   = "lca_labelHyper",
    number = Inf,
    sort.by = "none"
) |>
    as_tibble(rownames = "ensg")

cat("limma/voom DE genes (BH < 0.05):", sum(res_limma$adj.P.Val < 0.05), "\n")
```

limma/voom identifies **7,230 DE genes** at BH < 0.05: 3,368 upregulated
in Hyper and 3,862 upregulated in Hypo.

```{r limma-pval-hist}
res_limma |>
    filter(!is.na(P.Value)) |>
    ggplot(aes(x = P.Value)) +
    geom_histogram(binwidth = 0.02, boundary = 0,
                   fill = "steelblue", color = "white", alpha = 0.8) +
    geom_hline(
        yintercept = sum(!is.na(res_limma$P.Value)) * 0.02,
        linetype = "dashed", color = "firebrick"
    ) +
    labs(
        x        = "p-value",
        y        = "Count",
        title    = "limma/voom p-value distribution: Hyper vs Hypo",
        subtitle = "Dashed line = expected count per bin under global null"
    ) +
    theme_bw()
```

The limma/voom p-value distribution closely mirrors the DESeq2 distribution,
with a pronounced spike near zero consistent with widespread true signal.

```{r limma-plot-data}
limma_plot_df <- res_limma |>
    filter(!is.na(adj.P.Val)) |>
    mutate(label = case_when(
        adj.P.Val < 0.05 & logFC > 0 ~ "Up in Hyper",
        adj.P.Val < 0.05 & logFC < 0 ~ "Up in Hypo",
        TRUE                          ~ "NS"
    ))
```

```{r limma-volcano}
limma_plot_df |>
    ggplot(aes(x = logFC, y = -log10(adj.P.Val), color = label)) +
    geom_point(alpha = 0.4, size = 0.8) +
    scale_color_manual(values = c(
        "Up in Hyper" = "firebrick",
        "Up in Hypo"  = "steelblue",
        "NS"          = "grey70"
    )) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey40") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    labs(
        x        = "log2 fold change (Hyper / Hypo)",
        y        = "-log10(adjusted p-value)",
        color    = NULL,
        title    = "limma/voom volcano: Hyper vs Hypo",
        subtitle = "logFC from eBayes fit | BH-adjusted p-value"
    ) +
    theme_bw()
```

The limma/voom volcano is broadly symmetric, with Hypo-upregulated
genes (3,862) slightly outnumbering Hyper-upregulated genes (3,368),
a modest directional reversal compared to the DESeq2 result,
reflecting differences in variance estimation between the two methods.

```{r limma-ma}
limma_plot_df |>
    ggplot(aes(x = AveExpr, y = logFC, color = label)) +
    geom_point(alpha = 0.4, size = 0.8) +
    scale_color_manual(values = c(
        "Up in Hyper" = "firebrick",
        "Up in Hypo"  = "steelblue",
        "NS"          = "grey70"
    )) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    labs(
        x        = "Average log2-CPM",
        y        = "log2 fold change (Hyper / Hypo)",
        color    = NULL,
        title    = "limma/voom MA plot: Hyper vs Hypo"
    ) +
    theme_bw()
```

The MA plot shows good centering of fold changes around zero at high
average expression, with the expected increase in LFC spread at lower
expression levels.

---

### Method comparison

```{r method-comparison}
# Join on shared genes, using unshrunken DESeq2 LFC for a fair comparison
comparison_df <- inner_join(
    as_tibble(res_deseq2, rownames = "ensg") |>
        dplyr::select(ensg, deseq2_lfc = log2FoldChange,
                      deseq2_pval = pvalue, deseq2_padj = padj),
    res_limma |>
        dplyr::select(ensg, limma_lfc = logFC,
                      limma_pval = P.Value, limma_padj = adj.P.Val),
    by = "ensg"
) |>
    filter(!is.na(deseq2_padj), !is.na(limma_padj)) |>
    mutate(
        sig_deseq2 = deseq2_padj < 0.05,
        sig_limma  = limma_padj  < 0.05,
        concordance = case_when(
            sig_deseq2 & sig_limma  ~ "Both",
            sig_deseq2 & !sig_limma ~ "DESeq2 only",
            !sig_deseq2 & sig_limma ~ "limma only",
            TRUE                    ~ "Neither"
        )
    )

# Counts per concordance category
comparison_df |> dplyr::count(concordance, sort = TRUE)
```

6,544 genes are significant in both methods, while 731 are
DESeq2-specific and 686 are limma-specific — representing high
concordance given the independent estimation frameworks.
The 8,496 genes non-significant in both form the majority of
the tested transcriptome.

```{r lfc-correlation}
# LFC concordance between methods
comparison_df |>
    slice_sample(n = min(5000, nrow(comparison_df))) |>
    ggplot(aes(x = deseq2_lfc, y = limma_lfc, color = concordance)) +
    geom_point(alpha = 0.3, size = 0.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
    scale_color_manual(values = c(
        "Both"        = "purple",
        "DESeq2 only" = "firebrick",
        "limma only"  = "steelblue",
        "Neither"     = "grey70"
    )) +
    labs(
        x        = "DESeq2 log2FC",
        y        = "limma/voom log2FC",
        color    = NULL,
        title    = "LFC concordance: DESeq2 vs limma/voom",
        subtitle = "Dashed line = identity"
    ) +
    theme_bw()
```

LFC estimates are highly concordant between methods (Spearman ρ = 0.94),
with points closely tracking the identity line across the full fold-change
range.

```{r pval-correlation}
# P-value concordance: -log10 scale
comparison_df |>
    slice_sample(n = min(5000, nrow(comparison_df))) |>
    ggplot(aes(x = -log10(deseq2_pval), y = -log10(limma_pval),
               color = concordance)) +
    geom_point(alpha = 0.3, size = 0.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
    scale_color_manual(values = c(
        "Both"        = "purple",
        "DESeq2 only" = "firebrick",
        "limma only"  = "steelblue",
        "Neither"     = "grey70"
    )) +
    labs(
        x        = "-log10(p): DESeq2",
        y        = "-log10(p): limma/voom",
        color    = NULL,
        title    = "P-value concordance: DESeq2 vs limma/voom",
        subtitle = "limma stronger among weak-evidence genes; DESeq2 dominates the significant tail"
    ) +
    theme_bw()
```

```{r pval-direction-table}
# Fraction of genes where each method assigns stronger evidence, by -log10(p) bin
tmp_dir <- comparison_df[!is.na(comparison_df$deseq2_pval) & !is.na(comparison_df$limma_pval), ]
tmp_dir$d   <- -log10(tmp_dir$deseq2_pval)
tmp_dir$l   <- -log10(tmp_dir$limma_pval)
tmp_dir$direction <- ifelse(tmp_dir$l > tmp_dir$d, "limma stronger", "DESeq2 stronger")
tmp_dir$bin <- cut(tmp_dir$d,
                   breaks = c(0, 1, 2, 5, 10, Inf),
                   labels = c("[0, 1)", "[1, 2)", "[2, 5)", "[5, 10)", "≥10"),
                   right  = FALSE)

tbl_dir <- as.data.frame.matrix(table(tmp_dir$bin, tmp_dir$direction))
tbl_dir$Total           <- rowSums(tbl_dir)
tbl_dir$`% limma stronger` <- round(100 * tbl_dir[["limma stronger"]] / tbl_dir$Total, 1)
tbl_dir$`DESeq2 -log10(p)` <- rownames(tbl_dir)
knitr::kable(
    tbl_dir[, c("DESeq2 -log10(p)", "DESeq2 stronger", "limma stronger", "Total", "% limma stronger")],
    row.names = FALSE,
    caption   = "Direction of evidence advantage by DESeq2 significance bin"
)
```

Evidence strength is similarly well-correlated overall (Spearman ρ = 0.91 on
the −log10(p) scale), but the direction of the advantage is not uniform. Limma
tends to assign higher significance at the lower end of the spectrum, while 
DESeq2 tends to assign lower pvalues to highly significant genes

```{r upset-overlap}
# Concordance counts split by effect direction for jointly significant genes
comparison_df$dir_concordance <- with(comparison_df, {
    ifelse(concordance == "Both" & deseq2_lfc > 0, "Both (up)",
    ifelse(concordance == "Both" & deseq2_lfc < 0, "Both (down)",
           concordance))
})
comparison_df$dir_concordance <- factor(comparison_df$dir_concordance,
    levels = c("Both (up)", "Both (down)", "DESeq2 only", "limma only", "Neither"))

dir_counts <- as.data.frame(table(comparison_df$dir_concordance))
names(dir_counts) <- c("dir_concordance", "n")

ggplot(dir_counts, aes(x = dir_concordance, y = n, fill = dir_concordance)) +
    geom_col(alpha = 0.85, show.legend = FALSE) +
    scale_fill_manual(values = c(
        "Both (up)"   = "#7B2D8B",
        "Both (down)" = "#C299CC",
        "DESeq2 only" = "firebrick",
        "limma only"  = "steelblue",
        "Neither"     = "grey70"
    )) +
    geom_text(aes(label = scales::comma(n)), vjust = -0.4, size = 3.2) +
    scale_y_continuous(labels = scales::comma,
                       expand = expansion(mult = c(0, 0.08))) +
    labs(x = NULL, y = "Number of genes",
         title = "DE gene overlap by direction: DESeq2 vs limma/voom",
         subtitle = "BH-adjusted p < 0.05; direction = sign of log2FC") +
    theme_bw()
```

The 6,544 jointly significant genes represent ~90% of all DESeq2 hits and
~90% of all limma hits. Among these, every gene with a concordant call also
has a concordant direction of effect — 3,334 upregulated and 3,210
downregulated in Hyper vs. Hypo — with zero direction conflicts.
The method-specific hits (731 DESeq2-only, 686 limma-only) are the only region
of disagreement, and these are concentrated among lower-evidence
genes as described above.

```{r}
# write_csv(comparison_df, here("data/earli_limma_deseq_comp.csv"))
```

